\documentclass[12pt]{article}
\usepackage[a4paper, margin=0.8in]{geometry}
\pdfpagewidth 8.5in
\pdfpageheight 11.0in
\textheight 9.7in
\usepackage{color} %used for font color
\usepackage{amssymb, amsmath, multirow, commath, inputenc, titling, epic, paralist, graphicx, algorithm,algorithmic, tikz, xcolor,colortbl, environ, quoting, ragged2e, lipsum}

\usepackage{sectsty}
\usepackage{multicol}
\sectionfont{\large}

\setlength{\droptitle}{-5.5em}
\setlength{\parindent}{30pt}

\begin{document}

\title{{\bf \large Utilizing an AI Approach to Pathfinding in Super Mario}\vspace{-2ex}}
\author{{\small Nick De Tullio, Michael Jalkio, \& Thomas Gautier}
\\ {\small nrd24@cornell.edu, mrj77@cornell.edu, tng26@cornell.edu}\vspace{-9ex}}
\date{}

\maketitle

\section * {Abstract}
This paper utilizes the Mario AI benchmark, to describe the success of both previously implemented Mario bots compared to our own implementation. The Mario AI Benchmark was introduced for use first in the IEEE Games Innovation Conference (ICE-GIC) and then later in Computational Intelligence Games (CIG) as a benchmark for reinforcement learning algorithms and other game AI techniques in a public domain clone of the platform game {\it Super Mario Bros}. Each year the CIG conducts a Mario AI Competition, in which they compare contestants submitted Mario bots to this benchmark. The goal of this paper is to determine the success of bots that were already created for this competition and then to decide whether an improved bot can be created. The bots that will be evaluated fell into three categories in the competitions A*-based, learning-based, and rule-based implementations.

\setlength{\columnsep}{.65cm}
\begin{multicols}{2}
\section * {Introduction}
A platform game, or platformer, is a video game which involves guiding an avatar to jump between suspended platforms, over obstacles, or both in order to advance the game. These challenges are known as jumping puzzles or freerunning. It is the job of the player to control the jumps in order to avoid letting their avatar fall from the platforms or miss the necessary jumps. The most common element of the genre of platform games is the jump button. Platform games originated in the early 1980s in side scrolling or 2D video games, which were eventually followed by 3D successors in the mid-1990s.
\\
\indent Super Mario Bros is a series of platform video games created by Nintendo, featuring the main character Mario who serves as the player's avatar. The game follows Mario's adventures in the fictional Mushroom Kingdom, where he runs and jumps across platforms and atop enemies in themed levels. In the game it is necessary for Mario to jump over enemies and between platforms. It is also beneficial to Mario to jump on enemies in order to prevent them from coming across his path again, provided that they do not have spiked turtle shells. Aside from staying alive and progressing towards the finish line, it is also beneficial to the player's score for Mario to collect coins by jumping into them as they are found in the level, as well as by jumping into bricks for coins that may be hidden. There are also a multitude of power-ups and items which give Mario special powers such as fireball-throwing, size-changing, and extra lives, which can also be found by jumping into bricks.
\\
\indent In order to make the benchmark possible the game was modified via the construction of an API that enabled it to be easily interfaced with learning algorithms and the various competitors' controllers. The modifications included the removal of the dependency on the system clock so that the learning algorithm can ``step" forward, removing the dependency on the graphical output, and substantial refactoring. Each step in the game corresponds to 40 milliseconds of simulated time which has an update frequency of 25 frames per second. At each step, the controller receives a description of the environment, and outputs an action. The software that makes these modifications possible is a single threaded Java application, which allows for the key methods that a controller needs to implement to be specified in a single Java interface file.
\\
\indent There are many features that make \textit {Super Mario Bros} or platformers in general particularly interesting from an artificial intelligence or reinforcement learning perspective. What is likely the most important feature of the game is its potentially very rich and high-dimensional environment representation. When a human player views the game, he views a small part of the current level in two dimensions, with the screen centered on Mario. There are very rarely sparse views, in most cases there are dozens of objects such as brick blocks, enemies and collectable items. The static environment, such as grass, pipes, and brick blocks are laid out in a grid that covers approximately $19 * 19$ cells.

\end{multicols}

\end{document}